---
title: "Gov 50 Final Project"
author: "Samantha O'Connell"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---



## Project thoughts

I am interested in exploring data related to the criminal justice system and wrongful convictions. The Death Penalty Information Center, an organization I've interned for in the past, recently released their "Death Penalty Census," a thorough tracking of all death sentences handed down in the US since 1972, and I would be particularly interested in interacting with this data. DPIC also has tracked every single death-row exoneration we have record of and I think this information would be a useful starting point for my inquiries into demographics about the defendants and victims and the states and counties in cases that were tainted by wrongful convictions. In light of the Biden administration's pardoning of all federal simple marijuana convictions, I'd be interested in tracking data about federal drug convictions in general, tracking different sentencing outcomes and demographics to see the potential disparities.


## Research Proposal

How does a capital defendant's race impact the eventual outcome of their death sentence in the United States? The history of capital punishment in the US is one deeply entwined with the legacy of slavery, white supremacy, and lynchings. Like the prison system as a whole, Black men are disproportionately sentenced to death, particularly in the South, and many exonerations in recent history have demonstrated repeated patterns of prosecutorial misconduct and unfair  jury selection. The Death Penalty Information Center, a nonprofit research think tank dedicated to exposing the ugly truths of the death penalty in American through first-rate statistical research, just recently released the Death Penalty Census (accessed here: https://deathpenaltyinfo.org/facts-and-research/death-penalty-census) It is the most thorough tracking of every capital sentence handed down since 1974 ever compiled, tracking important factors such as defendant's race, state, age of defendant, outcome of sentence. Now that this wealth of previously unorganized data has been released, I want to investigate the average treatment effect of race on whether the death sentence is resentenced to life in prison, overturned completely, still pending, or resulted in an actual execution. Based on the pervasive racial biases within the justice system especially during the "War on Crime" throughout the 1970s-1990s, I predict that a higher proportion of sentences for white defendants will be changed to life in prison and that a higher proportion of sentences for Black defendants will end in an actual execution. DPIC's database tracks the race of each defendant, my explanatory variable, grouping it into the categories White, Black, Latinx, Asian, American Indian or Alaska Native, Pacific Islander, Other Race, and Unknown. If the defendant is multiracial, races are separated by a comma. The current case status, my outcome variable of interest, is also tracked in detail by DPIC, grouped in many different categories based on whether the sentence was reduced, commuted, overturned, still active, if defendant was executed or if they died on death row naturally instead. If the percentage of cases for White defendants that ends in resentencing is much lower than the same calculation for Black defendants, that would provide serious evidence to challenge my hypothesis. 

```{r}
library(tidyverse)
library(ggplot2)
library(infer)
```


```{r}
deathsentence <- read.csv("Death-Sentences-in-the-United-States-Since-1972-as-of-January-1-2021.csv")



time_graph <- deathsentence |>
  group_by(Year)|>
  summarize(n=n())|>
  ggplot(mapping = aes(x = Year, y = n)) +
  geom_line() +
  labs (x = "Year", y = "Number of Death Sentences",
        title = "Change in the Use of the Capital Punishment Over Time (1972-2021)")

time_graph

region_graph <- deathsentence |>
  group_by(Region) |>
  filter(County.and.State != "Military") |>
  summarize(n=n()) |>
  ggplot(mapping = aes(x=Region, y = n))+
  geom_col() +
  labs(x = "US Region", "Number of Death Sentences Since 1972")

region_graph
```

```{r}
deathsentence<-deathsentence|>
  mutate(executed = if_else(Outcome.of.Sentence == "Executed" | Outcome.of.Sentence == "Executed by a Different State" | Outcome.of.Sentence == "Executed for a Different Crime" , 1, 0 )) |>
  mutate(exonerated = if_else(Outcome.of.Sentence == "Exonerated" | Outcome.of.Sentence == "Grant of Relief (Never
Retried)", 1, 0),
         minority = if_else(Races == "Black" | Races == "Latinx" | Races == "American Indian or Alaska Native" | Races == "Asian" | Races == "Other", 1, 0))|>
  mutate(Southern = if_else(Region == "South", 1, 0))


race_region_graph <- deathsentence |>
  group_by(Races, Region)|>
  filter(County.and.State != "Military",
         Races != "Unknown") |>
  summarize(
            meanexon = mean(exonerated)) |>
  ggplot(mapping = aes(x = Races, y = meanexon)) +
  geom_col() +
  facet_wrap(~ Region) +
  scale_x_discrete(labels = c("Native American", "Asian", "Black", "Latine", "Middle Eastern", "White"), guide = guide_axis(n.dodge=2)) +
  labs(x = "Race of Death Row Prisoner", y= "Proportion of Death Row Prisoners Ultimately Exonerated", title = "Rate of Capital Exonerations in Relation to Defendant's Race and US Region")

race_region_graph

fit_1 <- lm(exonerated ~ minority, data = deathsentence)
fit_1

fit_2 <- lm(exonerated ~ minority + Southern, data = deathsentence)
fit_2



fit_1 |>
  broom::tidy() |>
  select(term, estimate) |>  
  knitr::kable(digits = 3)

fit_2 |>
  broom::tidy() |>
  select(term, estimate) |>  
  knitr::kable(digits = 3)

ate<- deathsentence |>
  group_by(minority) |>
  summarize(mean = mean(exonerated, na.rm = TRUE))|>
pivot_wider(names_from = minority,
              values_from = mean) |>
  mutate(ATE = `1`-`0`)

knitr::kable(ate, col.names = c("Average Rate of Exoneration for White Defendants", "Rate of Exoneration for Nonwhite Candidates", "Difference in Means"),
             digits = 3)
```
## Analysis


As my visualizations all cumulatively demonstrate, the defendant being a minority seems to correlate with a higher rate of exoneration from death row. Basically there is a small but still positive relationship between being a nonwhite defendant and having your sentence be overturned based on evidence of innocence. The ATE based on my calculations seems to be 0.008, a very similar statistic to the coefficient I have caluclated using linear regression as well. Basically, when a defendant is white (when minority = 0 as opposed to 1), the rate of exoneration is about 0.014 or a little over 1% (the intercept that was predicted) while the changing of the defendant's race is correlated, on average, with a 0.8 percent increase. However, as my graph mapping the distribution of sentences by region demonstrates, the South has a disproportionately high rate of death sentences. Perhaps the region of the sentence is a confounder, correlating with a higher proportion of nonwhite defendants in general and some sort of structure state-specific issue within the justice system that causes more wrongful convictions and thus exonerations overall in the South, independent of the defendant's race. However, my second attempt at running a linear regression counteracts that argument. When a defendant is sentenced in a non-Southern state, nonwhite defendants are still being exonerated at a rate of 0.008 higher than white ones, an almost identical coefficient to the one I calculated before. On the other hand, when a defendant is white, being sentenced in a Southern state as opposed to in any other region actually correlates with a decreased rate of exoneration at -0.002. Basically, the question of whether the death sentence was handed down in the South does not appear to be a confounder witin this data, even if the highest percentage of death sentences are located there.  



```{r}


fit_3 <- lm(exonerated ~ minority + Year, data = deathsentence)
fit_3
  
race_dist <- deathsentence |>
  mutate(nonwhite = if_else(Races == "Black" | Races == "Latinx" | Races == "American Indian or Alaska Native" | Races == "Asian" | Races == "Other", "Nonwhite", "White"))|>
  group_by(nonwhite)|>
  summarize(count = n(), 
            exon = mean(exonerated))
race_dist
knitr::kable(race_dist, col.names = c("Race of Capital Defendant", "Total Death Sentences", "Rate of Exoneration"),
             digits = 3)

region_dist <- deathsentence |>
  group_by(Southern)|>
  summarize(count = n())
race_dist
region_dist
  




```

